Tutorial_1

### 2024-10-16 Tutorial_1a.Rmd
add Tutorial 1a and custom functions file

Load the functions. Make sure you specify the same levels and labels for factor condition in the function plot_eha (line 210) as above (line 88).

```{r load-functions}
# function censor creates censored observations and discrete response times (drt), given a user-defined censoring time (timeout) and bin width in ms
censor <- function(df, timeout, bin_width){
  if(!(timeout %% bin_width == 0)){
    return("The censoring time must be a multiple of the bin width!")
  }
  if(timeout < 0 | bin_width < 0){
    return("Both timeout and bin_width must be larger than 0!")
  }
  df %>% mutate(right_censored = 0,
                rtc = ifelse(rt > timeout, timeout, rt) %>% round(digits=2), # censored response times
                right_censored = ifelse(rtc == timeout,1,right_censored), # 1 = right censored observation, 0 = observed rt
                drt = ceiling(rtc/bin_width), # drt = discrete response time
                cens_time = timeout, bin_width = bin_width) # save both user-defined parameters for plotting
}

# function ptb creates a person-trial-bin oriented data set
ptb <- function(df){
  df %>% uncount(weights = drt) %>% # create drt rows per trial
         group_by(trial) %>% 
         mutate(period = 1:n()) %>% # create time bin (or time period) ranks within each trial
         mutate(event = if_else(period == max(period) & right_censored == 0, 1, 0)) %>% # event = 1 indicates response occurrence
         ungroup()
}

# function setup_lt sets up a life table for each level of condition (1 independent variable)
setup_lt <- function(ptb){
  ptb %>% mutate(event = str_c("event", event)) %>%
          group_by(condition,period) %>% 
          count(event) %>% 
          ungroup() %>% 
          pivot_wider(names_from = event,
                      values_from = n) %>% 
          mutate(event0 = ifelse(is.na(event0),0,event0), # replace NA with 0
                 event1 = ifelse(is.na(event1),0,event1),
                 risk_set = event0 + event1) %>% # define the risk set
          mutate(hazard = (event1 / risk_set) %>% round(digits = 3)) %>% # calculate hazard estimate
          mutate(se_haz = sqrt((hazard * (1 - hazard)) / risk_set) %>% round(digits = 4)) %>% # standard error for hazard
          group_by(condition) %>%
          mutate(survival = (cumprod(1-hazard)) %>% round(digits = 4), # calculate survival estimate
                 term     = (cumsum(hazard / (risk_set * (1 - hazard)))) %>% round(digits = 7), # intermediate calculation
                 se_surv  = (survival * sqrt(term)) %>% round(digits = 5)  ) %>% # Greenwood's (1926) approximation
          ungroup() 
}

# function calc_ca calculates the conditional accuracies 
calc_ca <- function(df){
  df %>% filter(right_censored==0) %>%
         group_by(condition,drt,cens_time,bin_width) %>%
         summarize(ca = mean(acc) %>% round(digits = 2),
                   n = n(),
                   .groups = 'drop') %>%
         ungroup() %>%
         mutate(period = drt,
                se_ca = sqrt((ca * (1-ca)) / n) %>% round(digits = 3)) %>%
         select(-drt)
}

# function join_lt_ca joins the conditional accuracies to the life tables
join_lt_ca <- function(df1,df2){df1 %>% left_join(df2, join_by(condition,period))}

# function extract_median is used to extract the S(t).50 quantiles (i.e., the estimated median RTs)
extract_median <- function(df){
  above_pct50 <- df %>% 
      group_by(condition) %>%
      filter(survival >= .5) %>% 
      slice(n()) # take last row
  below_pct50 <- df %>% 
      group_by(condition) %>%
      filter(survival < .5) %>% 
      slice(1) # take first row
  # pull period above
  period_above <- pull(above_pct50, period)
  # pull survivor function values
  survival_above <- pull(above_pct50, survival)
  survival_below <- pull(below_pct50, survival)
  # estimate median by interpolation
  median_period <- period_above+((survival_above-.5)/(survival_above-survival_below))*((period_above+1)-period_above)
}

# function plot_eha plots the hazard, survivor, and ca(t) functions, when there is one independent variable
plot_eha <- function(df,subj,haz_yaxis){ # set the upper limit of the y-axis for the hazard functions to haz_yaxis == 1 when plotting the data of individual subjects (see line 244)
  library(patchwork)
  cutoff <- df %>% pull(cens_time) %>% max(na.rm=T)
  binsize <- df %>% pull(bin_width) %>% max(na.rm=T)
  median_period <- extract_median(df)
  n_conditions <- nlevels(df$condition)
  data_median <- c()
  for(i in 1:n_conditions){
    data_median <- append(data_median, c(median_period[i], median_period[i]))
  }

  data_medians <- tibble(period= data_median,
                         survival = rep(c(.5, 0),n_conditions),
                         condition = rep(1:n_conditions, each=2))
# plot the hazard functions
p1 <- df %>% ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=hazard)) +
  geom_point(aes(y=hazard), size=1) + labs(color="Condition") +
  geom_linerange(aes(ymin=hazard-se_haz, ymax=hazard+se_haz), show.legend = F) +
  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits = c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,haz_yaxis)) +
  labs(x="", y="h(t)", title = paste("Subject ", subj)) +
  theme(legend.background = element_rect(fill = "transparent"),
        panel.grid = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(angle=90))
# plot the survivor functions
p2 <-df %>%
  ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=survival), show.legend = F) +
  geom_point(aes(y=survival), size=1, show.legend = F) +
  geom_linerange(aes(ymin=survival-se_surv, ymax=survival+se_surv), show.legend = F) +
  # add vertical lines at the median RTs in the plot of the survivor functions using geom_path(). Make sure you apply the same levels and labels for the factor condition as above on line 83!
  geom_path(aes(x=period, y=survival, color=factor(condition, levels =c(1,2,3),labels=c("blank","congruent","incongruent"))),
            data = data_medians, 
            linetype = 3, show.legend = F) +

  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits=c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="S(t)",
       colour="Condition") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90))
# plot the conditional accuracy functions
p3 <-df %>%
  ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=ca), show.legend = F) +
  geom_point(aes(y=ca), size=1, show.legend = F) +
  geom_linerange(aes(ymin=ca-se_ca, ymax=ca+se_ca), show.legend = F) +
  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits=c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="Time bin t's endpoint (ms)", y="ca(t)",
       colour="Condition") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90))

p1/p2/p3
}
```

Apply the functions

@@

# view plot (column 8) for subject 1. 
pluck(data_final, 8, 1)

@@

```{r aggregated-data-plot}
# create single subject number, and unique trial number
data_agg <- data_wr %>% mutate(oldtrial=trial,
                               oldpid=pid,
                               trial = 1:n(),
                               pid = 1)

# function plot_eha_agg plots the hazard, survivor, and ca(t) functions for aggregated data.
plot_eha_agg <- function(df,subj,haz_yaxis){ 
  library(patchwork)
  cutoff <- df %>% pull(cens_time) %>% max(na.rm=T)
  binsize <- df %>% pull(bin_width) %>% max(na.rm=T)
  median_period <- extract_median(df)
  n_conditions <- nlevels(df$condition)
  data_median <- c()
  for(i in 1:n_conditions){
    data_median <- append(data_median, c(median_period[i], median_period[i]))
  }

  data_medians <- tibble(period= data_median,
                         survival = rep(c(.5, 0),n_conditions),
                         condition = rep(1:n_conditions, each=2))
# plot the hazard functions
p1 <- df %>% ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=hazard)) +
  geom_point(aes(y=hazard), size=1) + labs(color="Condition") +
  geom_linerange(aes(ymin=hazard-se_haz, ymax=hazard+se_haz), show.legend = F) +
  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits = c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,haz_yaxis)) +
  labs(x="", y="h(t)", title = "Aggregated data (N=6)") +           # New info in title
  theme(legend.background = element_rect(fill = "transparent"),
        panel.grid = element_blank(),
        legend.position = "top")
# plot the survivor functions
p2 <-df %>%
  ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=survival), show.legend = F) +
  geom_point(aes(y=survival), size=1, show.legend = F) +
  geom_linerange(aes(ymin=survival-se_surv, ymax=survival+se_surv), show.legend = F) +
  geom_path(aes(x=period, y=survival, color=factor(condition, levels =c(1,2,3),labels=c("blank","congruent","incongruent"))),
            data = data_medians, 
            linetype = 3, show.legend = F) +

  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits=c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="S(t)",
       colour="Condition") +
  theme(panel.grid = element_blank())
# plot the conditional accuracy functions
p3 <-df %>%
  ggplot(aes(x=period, color=condition, group=condition)) +
  geom_line(aes(y=ca), show.legend = F) +
  geom_point(aes(y=ca), size=1, show.legend = F) +
  geom_linerange(aes(ymin=ca-se_ca, ymax=ca+se_ca), show.legend = F) +
  scale_x_continuous(breaks = c(0,1:(cutoff/binsize)), labels=c(0,1:(cutoff/binsize)*binsize),
                     limits=c(0,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="Time bin t's endpoint (ms)", y="ca(t)",
       colour="Condition") +
  theme(panel.grid = element_blank())

p1/p2/p3
}

# apply functions
data_agg_nested <- data_agg %>% group_nest(pid)
data_agg_final <- data_agg_nested %>% 
                     mutate(censored  = map(data, censor, 600, 40)) %>%   # ! user input: censoring time, and bin width
                     mutate(ptb_data  = map(censored, ptb)) %>%           # create person-trial-bin dataset
                     mutate(lifetable = map(ptb_data, setup_lt)) %>%      # create life tables without ca
                     mutate(condacc   = map(censored, calc_ca)) %>%       # calculate ca
                     mutate(lifetable_ca = map2(lifetable, condacc, join_lt_ca)) %>%    # create life tables with ca
                     mutate(plot      = map2(.x = lifetable_ca, .y = pid, plot_eha_agg,1))  # create plots 
# view plot
pluck(data_agg_final, 8, 1)
```

>>>>

Load our custom functions. Make sure you specify the same levels and labels for factor condition in the function plot_eha (line 118 in custom_functions.R) as above (line 88 in Tutorial_1a).

```{r load-functions}
source("custom_functions.R")
```

Apply the custom functions

@@

# view plot (column 8) for subject 1. 
pluck(data_final, 8, 6)

@@

```{r agg-data}
# create single subject number, and unique trial number
data_agg <- data_wr %>% mutate(oldtrial=trial,
                               oldpid=pid,
                               trial = 1:n(),
                               pid = 1)
```

```{r agg-descr}
# apply functions
data_agg_nested <- data_agg %>% group_nest(pid)
data_agg_final <- data_agg_nested %>% 
                     mutate(censored  = map(data, censor, 600, 40)) %>%   # ! user input: censoring time, and bin width
                     mutate(ptb_data  = map(censored, ptb)) %>%           # create person-trial-bin dataset
                     mutate(lifetable = map(ptb_data, setup_lt)) %>%      # create life tables without ca
                     mutate(condacc   = map(censored, calc_ca)) %>%       # calculate ca
                     mutate(lifetable_ca = map2(lifetable, condacc, join_lt_ca)) %>%    # create life tables with ca
                     mutate(plot      = map2(.x = lifetable_ca, .y = pid, plot_eha_agg,1))  # create plots 
```

```{r agg-plot}
# view plot
pluck(data_agg_final, 8, 1)
```
